---
title: "From tidyverse to python"
author: "Patrick Ward"
date: "3/29/2023"
output: html_document
---

## NOTE

This RMarkdown file is a companion for my **From tidyverse to python** `jupyter notebook`, which walks through the same steps using python.

## Import libraries

```{r}
library(tidyverse)
library(palmerpenguins)

theme_set(theme_classic())
```


## Load data set

```{r}
data("penguins")
df <- penguins

df %>%
  head()

```


## Exploratory Data Analysis

Count the number of species in the data

```{r}
df %>%
  count(species, sort = TRUE)
```


Count the number of islands in the data

```{r}
df %>%
  count(island, sort = TRUE)
```


Count the species on each island

```{r}
df %>%
  count(species, island, sort = TRUE)
```


Create a 3x3 table of species by island

```{r}
table(species = df$species,
      island = df$island)
```

Turn the 3x3 table into a table of proportions

```{r}
table(species = df$species,
      island = df$island) / nrow(df)
```


Barplot of species

```{r}
df %>%
  count(species, sort = TRUE) %>%
  ggplot(aes(x = reorder(species, -n), y = n)) +
  geom_col(aes(fill = species)) +
  xlab('species')
```


Describe the data in all columns

```{r}
psych::describe(df)
```


Calculate the mean for bill length and bill depth

```{r}
df %>%
  select(bill_length_mm, bill_depth_mm) %>%
  summarize(across(.cols = everything(),
                   ~mean(.x, na.rm = TRUE)))
```


Count the NAs in each column

```{r}
df %>%
  summarize(across(.cols = everything(),
                   ~sum(is.na(.x)))) %>%
  pivot_longer(cols = everything())
  
```


If we wanted to remove all rows with an NA we could run the following code. (We will comment it out for now).

```{r}
# df <- df %>% na.omit()
```


Get the 25th, 50th, and 75th quantiles for bill length and bill depth on the Island Biscoe


```{r}
df %>%
  filter(island == "Biscoe") %>%
  select(bill_length_mm, bill_depth_mm) %>%
  summarize(length25 = quantile(bill_length_mm, probs = 0.25, na.rm = TRUE),
            length50 = quantile(bill_length_mm, probs = 0.50, na.rm = TRUE),
            length75 = quantile(bill_length_mm, probs = 0.75, na.rm = TRUE),
            depth25 = quantile(bill_depth_mm, probs = 0.25, na.rm = TRUE),
            depth50 = quantile(bill_depth_mm, probs = 0.50, na.rm = TRUE),
            depth75 = quantile(bill_depth_mm, probs = 0.75, na.rm = TRUE))
```


Boxplot of bill length by species

```{r}
df %>%
  ggplot(aes(x = species, y = bill_length_mm, fill = species)) +
  geom_boxplot()
```


Make the boxplot horizontal

```{r}
df %>%
  ggplot(aes(x = bill_length_mm, y = species, fill = species)) +
  geom_boxplot()
```


Density plot of bill length grouped by sex

```{r}
df %>%
  na.omit() %>%
  ggplot(aes(x = bill_length_mm, fill = sex)) +
  geom_density(alpha = 0.4)
```


Bill Length by Sex with Facet's for Species

```{r}
df %>%
  na.omit() %>%
  ggplot(aes(x = bill_length_mm, fill = sex)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~species)
```


scatter plot showing flipper length relative to body mass

```{r}
df %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```


scatter plot with regression line

```{r}
df %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm")
```

scatter plot colored by sex

```{r}
df %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g,
             color = sex)) +
  geom_point()
```


scatter plot colored by sex with regression lines

```{r}
df %>%
  filter(!is.na(sex)) %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g,
             color = sex)) +
  geom_point() +
  geom_smooth(method = "lm")
```


Change in body mass across years by species

```{r}
df %>%
  group_by(species, year) %>%
  summarize(N = n(),
            avg = mean(body_mass_g, na.rm = TRUE),
            SE = sd(body_mass_g, na.rm = TRUE) / sqrt(N),
            .groups = "drop") %>%
  ggplot(aes(x = year, y = avg)) +
  geom_ribbon(aes(ymin = avg - SE, ymax = avg + SE,
                  fill = species),
              alpha = 0.3) +
  geom_line(aes(color = species))
```


## Group By Summarize

Mean and standard deviation of bill length for each species.

```{r}
df %>%
  group_by(species) %>%
  summarize(avg_bill_length = mean(bill_length_mm, na.rm = TRUE),
            sd_bill_length = sd(bill_length_mm, na.rm = TRUE))
```


Mean and standard deviation of bill length for each species by sex

```{r}
df %>%
  filter(!is.na(sex)) %>%
  group_by(species, sex) %>%
  summarize(avg_bill_length = mean(bill_length_mm, na.rm = TRUE),
            sd_bill_length = sd(bill_length_mm, na.rm = TRUE))
```


## Group By Mutate

Add a new column to the data frame that shows mean flipper length grouped by species

```{r}
df <- df %>%
  group_by(species) %>%
  mutate(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE))

df %>%
  head()
```


Z-score by species for flipper length 

```{r}
df <- df %>%
  group_by(species) %>%
  mutate(flipper_z = (flipper_length_mm - mean(flipper_length_mm, na.rm = TRUE)) / sd(flipper_length_mm, na.rm = TRUE))

df %>%
  head()
```


Add a column for the mean and standard deviation by species for bill length

```{r}
df <- df %>%
  group_by(species) %>%
  mutate(mean_val = mean(bill_length_mm, na.rm = TRUE),
         sd_val = sd(bill_length_mm, na.rm = TRUE))

df %>%
  head()
```


## ifelse / case_when

Create a new column that codes the islands as:

* Biscoe = Island1
* Dream = Island2
* Torgersen = Island3

```{r}
df <- df %>%
  mutate(island_code = case_when(island == "Biscoe" ~ "Island1",
                                 island == "Dream" ~ "Island2",
                                 island == "Torgersen" ~ "Island3"))

df %>%
  head()
```


## Linear Regression

Fit a linear model

```{r}
model <- lm(body_mass_g ~ flipper_length_mm + species + sex, data = df)
```


Review model results

```{r}
summary(model)
```

